Color- and piece-blind chess

### 2019-03-08 trunk/chess/papers/blind.tex
More actual progress here; need to rewrite the simul part now that I figured it out

@@

% (XXX So actually, simpler here is just: Loop until all white players have
% made their moves. Then copy all their boards to our board vector.)
While in State~\label{state:cmim1} (which we also start in), we
compare our stored source board states to the boards where we play as
black. Once the white opponent has made a move, resulting in $B'$,
there will be some position $B$ in our ``source'' set that is a valid
predecessor for $B'$ (that is, there exists some legal move $M$ that
advances $B$ to $B'$). For example, at the beginning, $B$ may be the
starting position (in fact it must, since all boards are initialized
this way) and $B'$ may be the position resulting from $1. e4$. With
the matching constructed (see below) we copy each $B'$ into our set of
boards, forgetting $B$.\footnote{In fact, we do not even need to
  construct the matching, although it may help us with bookkeeping
  (e.g., which games have already been copied?), depending on how the
  exhibition itself is represented.} Then we transition to
State~\label{state:cmim2}. In this state, we figure out what moves to
play in the pending boards. By invariant, each one $B$ will be
compatible with at least one of our source boards $B'$. Compatible
means that there exists a move $m$ where playing $m$ in board $B$
results in $B'$. If $m$ exists, it is unique (and easy to compute), so
this is the move that we will play. The only tricky thing is figuring
out which boards go with which; this is the problem of reconstructing
the perfect matching.

@@

Fortunately, we know the
 - there exists a perfect matching (players playing legally)
 - we can tell if we found one

so you have a set of black and white games, same number of each.
goal is to get an even score, while using O(n) space.
So we have an invariant that each of the games-as-white is paired
with at least one compatible game-as-black (and vice versa). Of
course at the beginning all pairs are compatible, and then white
makes a move; the games are still compatible because any of these
moves can be played on the black side. We just need to make sure
that we don't play e.g. e4 in all the games, because we need enough
black opponents to be playing against d4s and Nf3s and so on.
So, after all the white players have made their moves, we have
those boards, and need to make moves for black. For each, just pick
a compatible board (i.e., something that represents a legal move)
and play it. When we do so, we update the board to a state like
``I made this move as white.'' (So at the very beginning, all boards
are in the starting position with like ``I made this move as black,''
of course not really, more like ``white opponent to move'').
This is easy on the first move because all choices (that are in
the state ``copied this move from white'').

Now we wait for black players to make their moves. (Note waiting
can be done in constant space. You just compare each board that
has ``black opponent to move'' with the actual board state; if
the move is compatible then you update it. But might you confuse
the identities of two games? It would need to be the case that
there exists board states $B_1$, $B_2$, $B$ where $B$ can
legally follow $B_1$ and $B_2$ but $B_1$ and $B_2$ are different.
Of course this is possible! $B$ could be like

and then we have $B_1$ and $B_2$ as

\chessboard[tinyboard, setfen=1k6/8/8/8/8/8/8/K7 b - -, showmover]
\chessboard[tinyboard, setfen=Rk6/8/8/8/8/8/8/K7 b - -, showmover]

And so if we misassign $B$ to $B_1$, and then later see $B'$:

We are stuck because $B'$ cannot follow $B_2$.

So if we're actually going to do it with this method, we would
need to be able to compute the bipartite matching looking at all
the constraints. There certainly exists one, and any one will do. 
(Actually what we want is called a ``perfect matching.'')
So the question is, can we compute a perfect matching in constant
space? I mean, it seems like not. How could we even {\em store}
the matching? Well, we don't actually need to store the matching
because it's not *literally* what we need; we just want to have
assigned the games according to a perfect match. So one approach
is like, pick a random assignment. If we get stuck, just start
over (so we would have to keep two board states, in order to
be able to ``undo'', but that is still constant) and commit
the whole thing if we succeed. We're actually quite likely to
succeed since ambiguities are pretty rare. So this may work
fine in practice. The main thing we achieve here is that we
don't need to store the matching itself between rounds; the
matching is ``expensive'' because it contains an index, which
is logarithmic size in the number of games. So, with this
method, we get $O(n)$ space, which is actually around $200 n$.

For each pair, we store the state of the board $B$
as last we saw it in $G_w$ or $G_b$.

both Walice and Bob have made their moves (that is, it is white's turn
in $G_w$ and

$G_i$, we look up which pair $p$ it is, and look at $B_p$. This board

Note that

Computers can still impress us and each other with computational chess
feats other than playing. For example, computers are largely concerned
with filling up their memories with efficiently encoded data.

compactly storing a position. constant state

I think a version I liked was 64 bits (this gives the positions of all
pieces),\footnote{(We can also do a little better. Note that only up
  to 32 of the 64 bits can be set, so you could represent
  $\binom{60}{32}$ in $\sim 61$ bits and use e.g. a third king to
  indicate that there are no more actually occupied squares if fewer
  than 32 pieces are present. However, this gets vastly more
  complicated for only 3 bits of savings.)} followed by 32 records
giving the piece values (there are 6: pawn, knight, bishop, rook,
queen, king) and colors. To pick between the 12 possibilities we need
4 bits, and there is some slack here of course. anyway, that's 64 +
32*4 = 192 bits, plus 15 bits for other state. Note we could use the
c\_rook trick to save 4 bits for castling rights (14 possibilities),
not to mention a en\_passantable pawn (15) and a king whose turn it is
(even 16). Clear slack here, but that would leave only the 50-move
rule at 7 bits, for a total of 199. Wikipedia says the huffman
approach is ``maximum of 204 bits, and often much less.'' but the
article also contains many bugs, like the misconception that there can
only be four rooks.

So now comes the main idea of the paper, which is also spoilered in the
very clear paper title. What if we represented boards {\em only} as the
64-bit mask telling us what squares are occupied? The encoding is very
lossy, of course, but it often contains enough information to deduce
the state of the board, or ...

>>>>

% Computers can still impress us and each other with computational chess
% feats other than playing. For example, computers are largely concerned
% with filling up their memories with efficiently encoded data.

@@

% (XXX So actually, simpler here is just: Loop until all white players have
% made their moves. Then copy all their boards to our board vector.)
While in State~\label{state:cmim1} (which we also start in), we
compare our stored source board states to the boards where we play as
black. Once the white opponent has made a move, resulting in $B'$,
there will be some position $B$ in our ``source'' set that is a valid
predecessor for $B'$ (that is, there exists some legal move $M$ that
advances $B$ to $B'$). For example, at the beginning, $B$ may be the
starting position (in fact it must, since all boards are initialized
this way) and $B'$ may be the position resulting from $1. e4$. With
the matching constructed (see below) we copy each $B'$ into our set of
boards, forgetting $B$.\footnote{In fact, we do not even need to
  construct the matching, although it may help us with bookkeeping
  (e.g., which games have already been copied?), depending on how the
  exhibition itself is represented.} Then we transition to
State~\label{state:cmim2}. In this state, we figure out what moves to
play in the pending boards. By invariant, each one $B$ will be
compatible with at least one of our source boards $B'$; we can iterate
over source boards until we find one in constant space (and mark it
as ``taken'' with . Compatible
means that there exists a move $m$ where playing $m$ in board $B$
results in $B'$. If $m$ exists, it is unique (and easy to compute), so
this is the move that we will play. The only tricky thing is figuring
out which boards go with which; this is the problem of reconstructing
the perfect matching.

@@

Fortunately, we know that there exists a perfect matching (assuming
all players are playing legal moves) and we can tell if we found
one (i.e., we didn't get stuck). So, one strategy that works is
to choose randomly when there is some ambiguity, and start over
from the beginning if we ever get stuck. In practice this will
be pretty efficient, since convergent board states are unusual.
We only need a single pseudorandom pool for the entire process,
so it can be O(n) bits; this seems easily enough to generate all
possible permutations of $n/2$ items. Even $n^{2n}$ grows much
faster than $n!$. If we don't like the random approach, I believe
it is also possible to compute \verb+next_permutation+ in
constant space; so we can just explicitly try all orderings for $B'$
(this takes exponential time).

Once we have paired up each $B$ and $B'$, we simply compute the
move (which we now know exists) and play it in that game. We
then wait for the black opponents to play their moves, copy the
resulting board states into our vector and repeat the above
process (but flipping ``black'' and ``white'').

Although this is more involved than the previous approach, and may
take exponential time, it allows us to play against $n$ simultaneous
opponents using $O(n)$ space!\footnote{ 
%
If the games in the exhibition were presented to us in some way that
allowed us to visit them in the same order multiple times, we could
perhaps get away with {\em only} the random pool? XXX }
%
The actual space used per game is primarily the representation of
a chess position, plus a few bits for bookkeeping. So, representing
boards compactly gives us a way to increase the number of simultaneous
games we can play for a given storage size.

Mainly, we need to store the pieces and their locations. There are a
few other bits, like whose turn it is (1 bit), whether each player can
still castle king- and queen-side (4 bits), and whether and where an
en passant capture is possible (4 bits).\footnote{
%
Technically, we need to store a lot of additional information with the
board in order to completely implement the rules of
chess.\cite{fiderules} The trickiest of these involve the rules for
draw by repetition, which make reference to the history of the game
(See Footnote~1 in {\it Survival in chessland}\cite{survival} and seem
to require storing all previous board states. Fortunately, if we are
being this picky, then we also know that the length of a chess game is
bounded by a constant: Rule 9.6.2 ends the game in a draw if both
players have made 75 moves without a pawn move or
capture,\footnotemark\ so it suffices to store the $75*2$ most recent
(half-)moves. This sucks so most people don't do it (for example, FEN
notation only gives the number of such moves, and so cannot implement
the draw by repetition). On the other hand, if we insist, then this
may give us a simpler route to a constant-space exhibition, since the
$B \stepsto{m} B'$ relation is probably reversible with such
information.}\footnotetext{These are two types of moves that make it impossible
  to formally repeat a position that preceded them. Castling also has
  this property, but doesn't count because it is a ``secret move.''}

With 64 squares, six pieces for each of the two sides, plus the empty
square, a straightforward representation of the board uses $64*4=256$
bits. The Thursd'z Institute considered more compact
representations\cite{representations}; one nice choice works as follows:

Start with a single 64-bit mask which indicates, for each square on
the board, whether it contains any piece. Note that there can only be
up to 32 pieces on the board. To tell what these pieces are, we then
follow with 32 four-bit records; these indicate the piece's color and
type.\footnote{(Since only 32 of the 64 bits can be set, you could do
  slightly better by representing $\binom{60}{32}$ in $\sim 61$ bits.
  When fewer than 32 squares are occupied, we can use a record
  containing e.g. a third king (otherwise impossible) to indicate that
  we should ignore the remaining bits. However, this gets vastly more
  complicated for only 3 bits of savings.)} 
%
With the 9 bits of extra state above, this totals $64+32*4+9=201$
bits. There is some slack in the representation, because there are
only 12 actual possibilities for a piece but we can indicate 16 with 4
bits. It is great to abuse this slack to save bits; for example, we
can store a new type of rook that is still able to castle (it can only
be in the corners and thus also indicates its own color), eliminating
the 4 castling bits. We can similarly introduce an en-passantable
pawn, saving the 4 bits of en passant state; this piece can only be in
ranks 4 or 5, so it also indicates its color. We can also have a
``king whose turn it is'' for each side, saving the side-to-move bit.
This totals a nice round $64+32*4=192$ bits.\footnote{ Since this is
  SIGBOVIK, I am freed from the burden of comparing related work. I
  did however read the rather bad Wikipedia article on the
  topic\cite{wikipediaboard} which describes a Huffman-based encoding
  that uses a ``maximum of 204 bits, and often much less.'' This also
  includes a 7-bit 50-move counter (but you really need to implement a
  75-move counter; 50 moves only allow a player to {\em claim} a draw)
  so should be compared as $197$ bits. But the article also contains
  many bugs, like the misconception that there can only be four total
  rooks (pawns are allowed to promote to rook). So the approach
  described here is both {\em more efficient} and {\em more correct}.}
%
This would allow approximately an 11 billion-game simultaneous
exhibition in RAM on my desktop computer.

\medskip
So now comes the main idea of the paper, which is also spoilered in the
very clear paper title. What if we represented boards {\em only} as the
64-bit mask telling us what squares are occupied? The encoding is very
lossy, of course, but it often contains enough information to deduce
the state of the board. For example, can you tell what position this is?

\chessboard[setfen=8/8/8/8/8/8/8/8,
  showmover=false, 
  smallboard,  
  color=gray,
  strokeopacity=0,
  markstyle={[fill]circle},
  markfields={a1,b1,c1,d1,e1,f1,g1,h1,a2,b2,c2,d2,e4,f2,g2,h2,a7,b7,c7,d7,e7,f7,g7,h7,a8,b8,c8,d8,e8,f8,g8,h8
}]

Correct! It is

\begin{tabular}{p{1.85in}p{1.4in}}
\chessboard[setfen=nrbqkbrn/pppppppp/8/8/4P3/8/PPPP1PPP/NRBBNRQK b - - 34 19, showmover=false, smallboard] & {
\vspace{-1.65in} \small after 1. Nf3 Nf6 2. e4 Ng4 3. Bc4 Ne5 4. O-O Ng6
5. Kh1 Rg8 6. Nc3 Nh8 7. Qe2 Nc6 8. Rb1 Rb8
9. Nb5 Nb4 10. Rd1 Nd5 11. Qf1 Nb6 12. Qg1 Na8
13. Nbd4 Nb6 14. Rf1 Na8 15. Be2 Ng6 16. Bd1 Nh8
17. Nb3 Ng6 18. Ne1 Nh8 19. Na1}
\end{tabular} \vspace{1em}

So this is kind of like blindfold chess ... XXX
